---
---
title: "Employee Attrition Analysis"
output: html_document
editor_options:
  chunk_output_type: console
---
---

```{r}
# 1. Setup

library(tidyverse); library(caret); library(pROC); library(rpart); library(rpart.plot)
```


```{r}
# 2. Load and inspect
df <- read_csv("week13_employee.csv")
glimpse(df)
summary(df)
colSums(is.na(df))

```


```{r}
# 3. Clean
# Columns with only one value (likely FTE, StandardHours) carry no information
df %>% summarise(across(everything(), n_distinct)) %>%
  pivot_longer(everything()) %>% filter(value == 1)
df <- df %>% select(-FTE, -StandardHours)
df <- df %>%
  mutate(across(where(is.character), as.factor),
         Attrition = factor(Attrition, levels = c("No", "Yes")))
```





```{r}
# 4. Exploratory analysis
df %>% group_by(Department) %>%
  summarise(n = n(), attrition_rate = mean(Attrition == "Yes"))

ggplot(df, aes(JobRole, fill = Attrition)) +
  geom_bar(position = "fill") + coord_flip() +
  labs(y = "Proportion")

ggplot(df, aes(Attrition, YearsAtCompany, fill = Attrition)) + geom_boxplot()

```



```{r}
# 5. Hypothesis tests
t.test(Age ~ Attrition, data = df)
chisq.test(table(df$BusinessTravel, df$Attrition))
```


```{r}
# 6. Logistic regression
set.seed(123)
idx   <- createDataPartition(df$Attrition, p = 0.7, list = FALSE)
train <- df[idx, ]
test  <- df[-idx, ]

fit <- glm(Attrition ~ ., data = train, family = binomial)
summary(fit)
exp(cbind(OR = coef(fit), confint.default(fit)))  

# odds ratios
or_table <- broom::tidy(fit, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  transmute(variable = term,
            OR       = round(estimate, 3),
            CI_low   = round(conf.low, 3),
            CI_high  = round(conf.high, 3),
            p_value  = round(p.value, 4)) %>%
  arrange(p_value)

or_table

```




```{r}
# 7. Evaluate
prob <- predict(fit, test, type = "response")
pred <- factor(ifelse(prob > 0.5, "Yes", "No"), levels = c("No", "Yes"))
confusionMatrix(pred, test$Attrition, positive = "Yes")

roc_obj <- roc(test$Attrition, prob)
plot(roc_obj); auc(roc_obj)

# Lower cutoff catches more leavers (trade-off: more false alarms)
pred30 <- factor(ifelse(prob > 0.3, "Yes", "No"), levels = c("No", "Yes"))
confusionMatrix(pred30, test$Attrition, positive = "Yes")

# AUC on the training data, to compare with the test AUC
auc(roc(train$Attrition, predict(fit, train, type = "response")))
```




```{r}
# 8. Optional: decision tree for interpretability
tree <- rpart(Attrition ~ ., data = train, method = "class")
rpart.plot(tree)

tree_pred <- predict(tree, test, type = "class")
confusionMatrix(tree_pred, test$Attrition, positive = "Yes")

tree_prob <- predict(tree, test, type = "prob")[, "Yes"]
auc(roc(test$Attrition, tree_prob))

# Which variables matter most
tree$variable.importance

```

```{r}
# tidymodels-setup
install.packages(c("tidymodels", "skimr", "corrr", "vip", "glmnet", "ranger", "xgboost"))
```
```
```{r}

```

```{r}
# prep-checks
# Missing values and variable types
skim(df)

# Class imbalance
df %>% count(Attrition) %>% mutate(prop = round(n / sum(n), 3))

# Variables with very few distinct values
df %>%
  summarise(across(everything(), n_distinct)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_unique") %>%
  arrange(n_unique)

# Skewness of numeric variables
skew <- function(x) mean((x - mean(x))^3) / sd(x)^3
df %>%
  select(where(is.numeric)) %>%
  summarise(across(everything(), skew)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "skewness") %>%
  arrange(desc(abs(skewness)))

# Highly correlated numeric predictors
df %>%
  select(where(is.numeric)) %>%
  correlate() %>%
  stretch() %>%
  filter(abs(r) > 0.7, as.character(x) < as.character(y)) %>%
  arrange(desc(abs(r)))
```

```{r eda}
df %>%
  group_by(Department) %>%
  summarise(n = n(), attrition_rate = round(mean(Attrition == "Yes"), 3))

ggplot(df, aes(JobRole, fill = Attrition)) +
  geom_bar(position = "fill") +
  coord_flip() +
  labs(y = "Proportion")

num_vars <- df %>%
  select(where(is.numeric)) %>%
  select(where(~ n_distinct(.x) > 1)) %>%
  names()

df %>%
  select(Attrition, all_of(num_vars)) %>%
  pivot_longer(-Attrition) %>%
  ggplot(aes(Attrition, value, fill = Attrition)) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(~name, scales = "free_y")
```

```{r bivariate}
# Numeric variables: Wilcoxon test
map_dfr(num_vars, ~ tibble(
  variable = .x,
  p_value  = wilcox.test(df[[.x]] ~ df$Attrition)$p.value
)) %>%
  arrange(p_value)

# Categorical variables: chi-square test
cat_vars <- df %>% select(where(is.factor), -Attrition) %>% names()

map_dfr(cat_vars, ~ tibble(
  variable = .x,
  p_value  = suppressWarnings(chisq.test(table(df[[.x]], df$Attrition))$p.value)
)) %>%
  arrange(p_value)
```
```
```{r}
# split
set.seed(123)
split <- initial_split(df, prop = 0.8, strata = Attrition)
train <- training(split)
test  <- testing(split)

folds <- vfold_cv(train, v = 5, strata = Attrition)
```

```{r recipes}
rec_base <- recipe(Attrition ~ ., data = train) %>%
  step_zv(all_predictors()) %>%
  step_YeoJohnson(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_normalize(all_numeric_predictors())

rec_corr <- rec_base %>%
  step_corr(all_numeric_predictors(), threshold = 0.8)

rec_pca <- rec_base %>%
  step_pca(all_numeric_predictors(), threshold = 0.90)
```

```{r models}
logit_spec <- logistic_reg() %>%
  set_engine("glm")

glmnet_spec <- logistic_reg(penalty = tune(), mixture = tune()) %>%
  set_engine("glmnet")

rf_spec <- rand_forest(trees = 500, mtry = tune(), min_n = tune()) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")

xgb_spec <- boost_tree(trees = tune(), tree_depth = tune(), learn_rate = tune()) %>%
  set_engine("xgboost") %>%
  set_mode("classification")
```

```{r tuning}
# Takes several minutes
metrics <- metric_set(roc_auc, pr_auc, sens, spec, f_meas)

wf_set <- workflow_set(
  preproc = list(base = rec_base, corr = rec_corr, pca = rec_pca),
  models  = list(logit = logit_spec, glmnet = glmnet_spec,
                 rf = rf_spec, xgb = xgb_spec),
  cross = TRUE
)

res <- wf_set %>%
  workflow_map(
    "tune_grid",
    resamples = folds,
    grid      = 10,
    metrics   = metrics,
    control   = control_grid(save_workflow = TRUE),
    seed      = 123,
    verbose   = TRUE
  )
```

```{r compare}
autoplot(res, metric = "roc_auc")

rank_results(res, rank_metric = "roc_auc", select_best = TRUE) %>%
  filter(.metric == "roc_auc") %>%
  select(wflow_id, mean, std_err, rank)
```
```
```{r}
 best-model
best_id <- rank_results(res, rank_metric = "roc_auc", select_best = TRUE) %>%
  filter(.metric == "roc_auc") %>%
  slice(1) %>%
  pull(wflow_id)
best_id

best_params <- extract_workflow_set_result(res, best_id) %>%
  select_best(metric = "roc_auc")

final_wf <- extract_workflow(res, best_id) %>%
  finalize_workflow(best_params)

final_fit <- last_fit(final_wf, split, metrics = metrics)
collect_metrics(final_fit)

preds <- collect_predictions(final_fit)
conf_mat(preds, truth = Attrition, estimate = .pred_class)
roc_curve(preds, truth = Attrition, .pred_Yes) %>% autoplot()
```

```{r threshold}
preds_30 <- preds %>%
  mutate(pred_30 = factor(if_else(.pred_Yes > 0.30, "Yes", "No"),
                          levels = c("Yes", "No")))

conf_mat(preds_30, truth = Attrition, estimate = pred_30)
sens(preds_30, truth = Attrition, estimate = pred_30)
spec(preds_30, truth = Attrition, estimate = pred_30)
```

```{r fit-check}
# Underfitting / overfitting check
fit_train <- fit(final_wf, train)

train_pred <- bind_cols(train %>% select(Attrition),
                        predict(fit_train, train, type = "prob"))
train_auc <- roc_auc(train_pred, Attrition, .pred_Yes)$.estimate

cv_auc <- rank_results(res, rank_metric = "roc_auc", select_best = TRUE) %>%
  filter(wflow_id == best_id, .metric == "roc_auc") %>%
  pull(mean)

test_auc <- collect_metrics(final_fit) %>%
  filter(.metric == "roc_auc") %>%
  pull(.estimate)

tibble(train = train_auc, cross_validation = cv_auc, test = test_auc)
```

```{r importance}
final_fit %>% extract_fit_parsnip() %>% vip(num_features = 15)
```
```





