---
title: "Week 13-IBM Employee Attrition"
author: "Hammed"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

set.seed(123)
```

##Load the dataset
```{r}
employee <- read.csv("week13_employee.csv")

head(employee)
```

##Check the structure
```{r}
str(employee)
```

##Dataset dimensions
```{r}
dim(employee)
```

##Summary statistics
```{r}
summary(employee)
```

##Check for missing values
```{r}
colSums(is.na(employee))
```

##Check Attrition distribution
```{r}
table(employee$Attrition)
```

##Chart
```{r}
barplot(
  table(employee$Attrition),
  col = c("red","steelblue"),
  main = "Employee Attrition Distribution",
  xlab = "Attrition",
  ylab = "Count"
)
```


#Remove non-informative variables
```{r}
employee <- subset(
  employee,
  select = -c(FTE, StandardHours)
)

str(employee)
```


##Convert Attrition to a factor
```{r}
employee$Attrition <- factor(
  employee$Attrition,
  levels = c("Yes","No")
)

str(employee$Attrition)
```

##Age vs Attrition
```{r}
boxplot(
  Age ~ Attrition,
  data = employee,
  col = c("tomato","steelblue"),
  main = "Age by Attrition",
  xlab = "Attrition",
  ylab = "Age"
)
```

##Years at Company vs Attrition
```{r}
boxplot(
  YearsAtCompany ~ Attrition,
  data = employee,
  col = c("tomato","steelblue"),
  main = "Years at Company by Attrition",
  xlab = "Attrition",
  ylab = "Years at Company"
)
```
##Total Working Years vs Attrition
```{r}
boxplot(
  TotalWorkingYears ~ Attrition,
  data = employee,
  col = c("tomato","steelblue"),
  main = "Total Working Years by Attrition",
  xlab = "Attrition",
  ylab = "Years"
)
```
##Business Travel vs Attrition
```{r}
travel_table <- table(
  employee$BusinessTravel,
  employee$Attrition
)

barplot(
  prop.table(travel_table,1),
  beside = FALSE,
  col = c("steelblue","red"),
  legend = rownames(t(travel_table)),
  main = "Business Travel and Attrition"
)
```

##Job Role vs Attrition
```{r}
role_table <- table(
  employee$JobRole,
  employee$Attrition
)

barplot(
  prop.table(role_table,1),
  beside = FALSE,
  col = c("steelblue","red"),
  las = 2,
  main = "Job Role and Attrition"
)
```

##Correlation Analysis
```{r}
numeric_data <- employee[,sapply(employee,is.numeric)]

cor_matrix <- cor(numeric_data)

round(cor_matrix,2)
```

##Convert Character Variables to Factors
```{r}
employee[] <- lapply(employee, function(x){
  if(is.character(x)) factor(x) else x
})

str(employee)
```

### Create training and testing datasets
```{r}
# Create training and testing datasets

set.seed(123)

train_index <- sample(
  1:nrow(employee),
  0.8 * nrow(employee)
)

train <- employee[train_index, ]

test <- employee[-train_index, ]

dim(train)
dim(test)
```

##Logistic regression
```{r}
model_logit <- glm(
  Attrition ~ .,
  data = train,
  family = binomial
)

summary(model_logit)
```

##Prediciton on Data
```{r}
pred_prob <- predict(
  model_logit,
  newdata = test,
  type = "response"
)

pred_class <- ifelse(
  pred_prob > 0.5,
  "No",
  "Yes"
)

table(
  Actual = test$Attrition,
  Predicted = pred_class
)
```

##Accuracy
```{r}
mean(
  pred_class == test$Attrition
)
```

```{r}
summary(model_logit)
```

```{r}
table(
Actual = test$Attrition,
Predicted = pred_class
)
```




